### 2018-05-28 write-up/abstract.tex
Add Acknowledgements
Fix many small writing errors

\newpage
{\normalfont\Huge\sffamily\bfseries TO DO: Abstract}
\vspace{24pt}

In this thesis, I argue that linear types are an appropriate, \emph{type-based
formalism} for expressing aliasing, read/write permissions, memory allocation,
re-use and deallocation, first, in the context of the APIs of linear algebra
libraries and then in the context of matrix-expression compilation.

I show that framing the problem using linear types can help \emph{reduce bugs}
by making precise and explicit the informal, ad-hoc practices typically
employed by human experts and linear algebra \emph{compilers} and automate
checking them.

As evidence for this argument, I show non-trivial, yet readable, linear algebra
programs, that are safe (with respect to aliasing, read/write permissions,
memory re-use and deallocation) and explicit (with respect to memory
allocation) which (1) are more memory-efficient than equivalent programs
written using purely high-level linear algebra libraries and (2) perform just
as predictably as equivalent programs written using low-level linear-algebra
libraries. I also argue \emph{the experience} of writing such programs with
linear types is ``better'' in key-respects. In addition to all of this, I show
that it is possible to provide such features \emph{as a library} on top of
existing programming languages and linear algebra libraries.

\textbf{Summarise evaluation.}

\textbf{Summary future work.}

>>>>

\newpage
{\normalfont\Huge\sffamily\bfseries Abstract}
\vspace{24pt}

In this thesis, I argue that linear types are an appropriate, \emph{type-based
formalism} for expressing aliasing, read/write permissions, memory allocation,
re-use and deallocation, first, in the context of the APIs of linear algebra
libraries and then in the context of matrix expression compilation.

I show that framing the problem using linear types can \emph{reduce bugs}
by making precise and explicit, the informal, ad-hoc practices typically
employed by experts and matrix expression compilers \emph{and} automate
checking them.

As evidence for this argument, I show non-trivial, yet readable, linear algebra
programs, that are safe and explicit (with respect to aliasing, read/write
permissions, memory allocation, re-use and deallocation) which (1) are more
memory-efficient than equivalent programs written using high-level
linear algebra libraries and (2) perform just as predictably as equivalent
programs written using low-level linear algebra libraries. I also argue
\emph{the experience} of writing such programs with linear types is
qualitatively better in key respects. In addition to all of this, I show that
it is possible to provide such features \emph{as a library} on top of existing
programming languages and linear algebra libraries.

\vspace*{\fill}
{\normalfont\Huge\sffamily\bfseries Acknowledgements}

I would like to thank my family and closest friends for supporting me through
my time at Cambridge: you have been my strength when I have had none. I would
also like to thank my supervisors, Dr\@.~Neelankantan Krishnaswami and
Dr\@.~Stephen Dolan, for teaching me so much this year. I owe a debt of
gratitude to both the 2017 Part III Appeals Committee and those who wrote
letters in support of my appeal, for believing I deserved the chance to stay
on: never have I valued the privilege of learning what I enjoy and enjoying
what I learn more.  Lastly, I would like to thank the many staff and research
members of the Computer Lab who have been approachable, patient and attentive
during my time here.

### 2018-05-28 write-up/background.tex
Fix minor layout issues
Fix many small writing errors

\prechapter{%
    I will outline the concept of linear types and show how they can be used to
    solve the problems faced by programmers writing code using linear-algebra
    libraries. I will emphasise \emph{practical} and intuitive explanations of
    linear types to keep this thesis accessible to programmers as well as
    academics not familiar with type-theory and will give only a terse overview
    of the history and theory behind linear types for the interested reader to
    pursue further.
}%

\section{Tracking Resources with Linearity}
Familiar examples of using a type-system to express program invariants are
existential-types for abstraction and encapsulation, polymorphic types for
parametricity and composition (a.k.a generics). Less-known examples include
dependent-types (contracts or pre- and post-conditions). The advantages of
using a type-system to express program invariants are summarised by saying the
stronger the rules you follow, the better the guarantees you can get about your
program, \emph{before} you run it. At first, the rules seem restrictive, but
similar to how the rules of grammar, spelling and more generally writing help a
writer make it easier and clearer to communicate the ideas they wish to express,
so too do typing rules make it easier to communicate the intent and
assumptions under which a program is written. An added, but often overlooked
benefit is automated-checking: a programmer can boldly refactor in certain
ways and the compiler will \emph{assist} in ensuring the relevant invariants
the type-system enforces are updated and kept consistent by pointing-out
where they are violated.

Linear types are a way to help a programmer track and manage resources. In
practical programming terms, they enforce the restriction that a value may be
used exactly once\footnote{This definition may differ from more colloquial uses
in discussions surrounding \emph{substructural} type systems and/or Rust.}
(Figure~\ref{fig:example_rules} shows two example typing rules for a typical,
linearly typed language). While this restriction may seem limiting at first,
precisely these constraints can be used to express common invariants of the
programs written by programmers every day. For example: a file or a
socket, once opened \emph{must} closed; all memory that is manually allocated
\emph{must} be freed.  C++'s destructors and Rust's Drop-trait (and more
generally, its borrow-checker) attempt to enforce these constraints by
basically doing the same thing: any resource that has not been moved is
deallocated at the end of the current lexical scope. Notably, these languages
also permit aliasing, alongside rules enforcing when it is acceptable to do so.
On face value, the above one-line description of linear types prevents aliasing
or functions such as $\lambda x.\; (x, x)$; such features are still allowed
(albeit in a more restricted fashion) in a \emph{usable} linear type system
designed for working with linear-algebra libraries.

@@

Given this background, the most pertinent question at hand is: what problems do
linear-algebra library users (and writers) typically face? The answer to this
question depends on which of two buckets a programmer falls (or is forced by
domain) into. On one side, we have users of high-level linear-algebra libraries such
as Owl (for OCaml), Julia and Numpy (for Python); other the other, we have
users of more manual, lower-level libraries such as BLAS (Basic Linear Algebra
Subroutines) for languages like C++ and FORTRAN.\footnote{I am not including
Rust in this comparison because its linear-algebra libraries are under active
development and not as well-known/used. Later on, given that it is a language
with in-built support of substructural features to track resources, Rust will
be compared and contrasted with this project to evaluate the classic
(E)DSL-versus-language-feature debate as it applies to the domain of
linear-algebra libraries.} Most of what follows applies to \emph{dense}
linear-algebra computations rather than \emph{sparse} because memory
allocated for results typically depends on the sparsity of the inputs and
so is not immediately amenable to the techniques proposed in this thesis.

@@

In Figure~\ref{fig:mat_mul_copies}, we see that matrix-multiplication is fairly
trivial to write and execute in Numpy, Julia and Owl. Let us call this approach
\emph{value-semantic}, meaning that objects are \emph{values} just like
integers and floating-point numbers. This approach confers two key advantages to the
programmmer: it is easy to read (equational and algebraic expressions) and it
is easy to reason about (as one would with a mathematical formula). Although this
approach does permit \emph{aliasing}, the consequences are benign because the
result of any computation is a \emph{new} value, distinct from any used during
the calculation of that value.

@@

However, these advantages come with some costs: constantly producing new values
is wasteful on memory (although the example given in
Figure~\ref{fig:mat_mul_copies} is only a $2 \times 2$ matrix, many real-world
datasets can contain up to gigabytes of data). A complex expression may create
many short-lived temporaries which would need to be reclaimed by a
garbage-collector (see Figure~\ref{fig:mat_mul_owl}). Libraries taking a
\emph{value-semantic} approach offer a dichotomy for a user wishing to
implement a new algorithm: either use the existing and safe primitives to build
an easy to reason about but slower, more memory-intensive algorithm, or use
escape-hatches (typically provided by most libraries, which permit in-place
modification of objects) to build faster, and more memory-efficient algorithms
which are harder to reason about.

\subsection{IHNIWTLM}
The title of this subsection\footnote{I Have No Idea What Those Letters Mean.}
illustrates the problem with the C++/FORTRAN side: readability (equational
and algebraic expressions) and ease of reasoning is sacrificed at
the altar of performance and efficiency.

Although escape-hatches do exist in value-semantic libraries, their use is
discouraged. Systematic consideration of performance requires lowering the
level of abstraction a programmer is working on. At this level, several
factors such as memory layout, allocation, re-use as well as cache behaviour
and parallelism become apparent. Of these, \textbf{memory allocation and
re-use} are relevant to linear-types and this thesis.

@@

In Fortran (Figure~\ref{fig:fortran_blas}), data is typically allocated
statically (at compile time) so temporary storage for all intermediate values
must be managed by the programmer. While this approach leads to verbose and
less readable code, the explicitness is good for understanding the memory
concerns of the program, albeit at the expense of understanding what the
program is actually calculating.

On the other hand, C++ (with operator overloading) can end up looking fairly
readable. For safety and correctness, expressions are typically handled with
value-semantics. However, given \emph{extra} information about, aliasing
(Eigen, Figure~\ref{fig:cpp_eigen}) or usage of intermediate expressions
(uBLAS, Figure~\ref{fig:cpp_ublas}), the number of temporaries allocated can be
reduced and increased \emph{implicitly} to improve performance (remove
unnecessary allocations or re-calculations respectively). Further tricks to
improve performance include expression templates (building up an
expression-tree at compile time and then pattern-matching on it to produce
code) and lazy evaluation (only calculating a result when it is needed). These
will be discussed in more detail in Chapter~\ref{chap:related}.

It is important to note that should these annotations (in
Figures~\ref{fig:cpp_eigen} and~\ref{fig:cpp_ublas}) be wrong, the program's
behaviour is very likely to end up being undefined (like how \texttt{memcpy()}
for overlapping regions of memory is explicitly undefined in the POSIX and C
standards). Indeed, one of Fortran's strengths lies in assuming that references
cannot be aliased (with certain caveats) in more cases than C permits (this
informal, general statement comes with many nuances left for the interested
reader to pursue).

@@

Let us have a look at a few examples of functions we can write with linear
types.  We can define the canonical factorial function
(Figure~\ref{fig:ltfla_factorial}) and sum over an array
(Figure~\ref{fig:ltfla_sumarray}).

@@

The syntax is intended to resemble OCaml's, apart from the spurious
\ltfla{`!'}s found here and there (they are annotations to show that we can
use a value more than once). In Figure~\ref{fig:ltfla_sumarray}, we see
\ltfla{row} has type (\ltfla{'x arr}). More detailed explanations of
\emph{what and why} will be given in Chapter~\ref{chap:impl}, but for now, it
is enough to know it means we can only \emph{read} from \ltfla{row}, and not
write to it. If we did try to write to or free \ltfla{row}, say by adding
another line as shown in Figure~\ref{fig:ltfla_univ}, would get a helpful error
message, as shown.

\begin{figure}[p]
    \begin{minted}[linenos, firstnumber=7, fontsize=\small]{ocaml}
let row = row[i] := x1 in (* or *) let () = free row in
(* Could not show equality:                                        *)
(*     z arr                                                       *)
(* with                                                            *)
(*     'x arr                                                      *)
(*                                                                 *)
(* Var 'x is universally quantified                                *)
(* Are you trying to write to/free/unshare an array you don't own? *)
(* In test/examples/sum_array.lt, at line: 7 and column: 19        *)
    \end{minted}
    \caption{Attempting to write to or free a read only array in
    LT4LA.}\label{fig:ltfla_univ}
\end{figure}

Now suppose we were trying to square a matrix, using a `dgemm' like BLAS
routine (called `simple\_dgemm'\footnote{For the purposes of this example, I
assume `simple\_dgemm' has type \ltfla{'x . 'x mat --o 'y .  'y mat --o z mat
--o ('x mat * 'y mat) * z mat}.} in Figures~\ref{fig:ltfla_square}
and~\ref{fig:ltfla_free}) which takes two read-only matrices and a third matrix
it can write to and performs $C := AB + C$. How would we use such a routine to
square a matrix, assuming $C=0$ initially?  Surely this would break linearity,
since $A$ would be have to be given to the function twice, and we can only do
so once?

We can use a special primitive called \ltfla{share} to produce (more) read-only
\emph{aliases} of any matrix. We then pass these into the function, and it
works as expected. Once the squared matrix has been obtained, we may not want
the original any more, and thus decide to free it. Before we do so, we must
first \ltfla{unshare} any read-only aliases that exist -- in this context,
\ltfla{unshare} swallows them and gives back a single, read-write handle.

\begin{figure}[p]
    \begin{minted}[linenos, fontsize=\small]{ocaml}
let (a1, a2) = share _ a in
let ((a1, a2), c) = simple_dgemm _ a1 _ a2 c in
let a = unshare _ a1 a2 in
let () = free a in c
    \end{minted}
    \caption{Squaring a matrix in LT4LA (assuming \ltfla{c} is initially
        0).}\label{fig:ltfla_square}
\end{figure}

If we tried to free one of the read-only aliases instead of
\ltfla{unshare}-ing, then we would get the error shown in
Figure~\ref{fig:ltfla_free}. Briefly, a \ltfla{z arr} is a read-write array,
everything else (\ltfla{'x arr} or \ltfla{_ s arr}) is read-only. The types of
\ltfla{free}, \ltfla{share} and \ltfla{unshare} (see
Appendix~\ref{chap:primitives}) are set up so that you can only free something
when you have read-write access to it, guaranteed by linearity to be the only
name in scope with this capability (aliases can only be read-only). Conversely,
if we \emph{forgot} to \ltfla{free a}, we would also get \texttt{Variable a not
used} error.

@@

\begin{figure}[p]
    \begin{minted}[linenos, fontsize=\small]{ocaml}
let (a1, a2) = share _ a in
let ((a1, a2), c) = simple_dgemm _ a1 _ a2 c in
let () = free a1 in c
(* Error:                                                        *)
(* Could not show equality:                                      *)
(*     z arr                                                     *)
(* with                                                          *)
(*     z s arr                                                   *)
(*                                                               *)
(* Could not show z and z s are equal.                           *)
(* Are you trying to write to/free an array before unsharing it? *)
(* In test.lt, at line: 3 and column: 17                         *)
    \end{minted}
    \caption{Attempting to free a read-only alias of matrix.}\label{fig:ltfla_free}
\end{figure}

@@

More generally, by annotating variables in the context with their usage (when
implementing a type-checker for a linearly typed language), we can express the
rules of \emph{substructural} (including affine, relevant and ordered type
systems) under the more general framework of \emph{co-effects}~\cite{petricek}.

@@

I have given an \emph{intuitive} exposition of linear types with
\emph{fractional-capabilities}~\cite{boyland}, emphasising small, but
illustrative and practical code examples, leaving \emph{what} is going on and
\emph{why} it works as details for the next chapter. I have shown that it is
possible to solve the dichotomy of readability, ease of reasoning and safety
versus memory-efficiency by providing explicit and automatically checked
management of aliasing, read/write permissions, memory allocation, re-use and
deallocation with a \emph{type system}. This prevents a whole class of errors
that can occur with lower-level languages \emph{at compile time}. I
demonstrated these features using the context of linear-algebra libraries as as
a specific example. In the next chapter, I will show that these features can be
further applied to the problem of \emph{matrix expression compilation}. I will
also explain how to express and implement them so that they can be provided and
used \emph{as a library}.

>>>>

\prechapter{%
    I will outline the concept of linear types and show how they can be used to
    solve the problems faced by programmers writing code that uses linear algebra
    libraries. I will emphasise \emph{practical} and intuitive explanations of
    linear types to keep this thesis accessible to programmers as well as
    academics not familiar with type theory and will give only a terse overview
    of the history and theory behind linear types for the interested reader to
    pursue further.
}%

\section{Tracking Resources with Linearity} Familiar examples of using a type
system to express program invariants are existential types for abstraction and
encapsulation, and polymorphic types for parametricity and composition (a.k.a
generics). Less-known examples include dependent types for contracts or pre-
and post-conditions. The advantages of using a type system to express program
invariants are summarised by saying: the stronger the rules you follow, the
better the guarantees you get about your program \emph{before} you run it. At
first, the rules may seem restrictive, but similar to how the rules of grammar,
spelling and more generally writing help a writer communicate their ideas more
clearly, so too do typing rules make it easier for the programmer to
communicate the intent and assumptions of their programs. An added, but often
overlooked benefit is automated-checking: a programmer can boldly refactor
their programs and the type checker will \emph{assist} in enforcing the
invariants it was given by highlighting where they were violated.

Linear types are a way to help a programmer track and manage resources. In
practical programming terms, they enforce the restriction that a value may be
used exactly once\footnote{This definition may differ from more colloquial uses
in discussions surrounding \emph{substructural} type systems and/or Rust.}
(Figure~\ref{fig:example_rules} shows two example typing rules for a typical,
linearly typed language). While this restriction may seem limiting at first,
precisely this constraint can be used to express common invariants of typical
programs. For example: a file or a socket, once opened \emph{must} closed; all
memory that is manually allocated \emph{must} be freed.  C++'s destructors and
Rust's Drop-trait (and more generally, its borrow-checker) attempt to enforce
these constraints by basically doing the same thing: any resource that has not
been moved is deallocated at the end of the current lexical scope. Notably,
these languages also permit aliasing, alongside rules enforcing when it is
acceptable to do so.  On face value, the above one-line description of linear
types prevents aliasing or functions such as $\lambda x.\; (x, x)$; such
features are still allowed (albeit in a more restricted fashion) in a
\emph{usable} linear type system designed for working with linear algebra
libraries.

@@

Given this background, the most pertinent question at hand is: what problems do
linear algebra library users (and implementors) typically face? The answer to
this question depends on which of two buckets a programmer falls (or is forced
by domain) into. On one side, we have users of high-level frameworks such as
Owl (for OCaml), Julia and Numpy (for Python); on the other, we have users
of more manual, low-level libraries such as BLAS (Basic Linear Algebra
Subroutines) for languages like C, C++ and Fortran.\footnote{I am not including
Rust in this comparison because its linear algebra libraries are under
active development and not as well-known/used. Later on, I will compare and
contrast Rust (a language with in-built support of substructural features
to track resources) with this project to evaluate the classic
(E)DSL-versus-language-feature debate as it applies to linear types.} Most of
what follows applies to \emph{dense} linear algebra computations rather than
\emph{sparse} ones because for the latter, memory allocated for results depends
on the sparsity of the inputs and so is not immediately amenable to the
techniques proposed in this thesis.

@@

In Figure~\ref{fig:mat_mul_copies}, we see that matrix multiplication is fairly
trivial to write and execute in Numpy, Julia and Owl. Let us call this approach
\emph{value-semantic}, meaning that objects are \emph{values} just like
integers and floating-point numbers. This approach confers two key advantages to the
programmmer: it is easy to read (equational and algebraic expressions) and it
is easy to reason about (as one would with a mathematical formula). Although this
approach does permit \emph{aliasing}, the consequences are benign because the
result of any computation is a \emph{new} value, distinct from any used during
the calculation of that value.

@@

However, these advantages come with some costs: constantly producing new values
is wasteful on memory (although the example given in
Figure~\ref{fig:mat_mul_copies} is only a $2 \times 2$ matrix, many real-world
datasets can contain up to gigabytes of data). A complex expression may create
many short-lived temporaries which would need to be reclaimed by a
garbage-collector (see Figure~\ref{fig:mat_mul_owl}). Libraries taking a
\emph{value-semantic} approach offer a dichotomy for a user wishing to
implement a new algorithm: either use the existing and safe primitives to build
an easy to reason about but slower, more memory-intensive algorithm, or use
escape-hatches (typically provided by most libraries, which permit in-place
modification of objects) to build faster, less readable and more
memory-efficient algorithms which are harder to reason about.

\subsection{IHNIWTLM}

The title of this subsection\footnote{I Have No Idea What Those Letters Mean.}
illustrates the problem with low-level libraries: readability (equational and
algebraic expressions) and ease of reasoning is sacrificed at the altar of
performance and (memory) efficiency.

Although escape-hatches do exist in high-level libraries, their use is
discouraged. Systematic consideration of performance requires lowering the
level of abstraction a programmer is working on. At this level, \textbf{several}
factors such as memory layout, allocation, re-use as well as cache behaviour
and parallelism are important. Of these, \textbf{only memory allocation and
re-use} are relevant to linear types and this thesis.

@@

In Fortran (Figure~\ref{fig:fortran_blas}) and C, temporary storage for all
intermediate values must be managed by the programmer. While this approach
leads to verbose, less readable and harder to reason about code, the
explicitness is good for understanding the memory concerns of the program.

On the other hand, C++ (with operator overloading) can end up looking fairly
readable. For safety and correctness, expressions are typically handled with
value-semantics (the result of any computation is a \emph{new} value, distinct
from any used during the calculation of that value). However, given
\emph{extra} information about aliasing (Eigen, Figure~\ref{fig:cpp_eigen}) or
usage of intermediate expressions (uBLAS, Figure~\ref{fig:cpp_ublas}), the
number of temporaries allocated can be reduced and increased \emph{implicitly}
to improve performance (remove unnecessary allocations or re-calculations
respectively). Further tricks to improve performance include expression
templates (building up an expression-tree at compile time and then
pattern-matching on it to produce code) and lazy evaluation (only calculating a
result when it is needed). These will be discussed in more detail in
Chapter~\ref{chap:related}.

It is important to note that should a \texttt{noalias} annotation be wrong, the
program's behaviour is very likely to end up being undefined (like how
\texttt{memcpy()} for overlapping regions of memory is explicitly undefined in
the POSIX and C standards). Indeed, one of Fortran's strengths lies in assuming
that references cannot be aliased (with certain caveats) in more cases than C
permits (this informal, general statement comes with many nuances left for the
interested reader to pursue).

@@

Let us have a look at a few examples of functions we can write with linear
types.  We can define the factorial function
(Figure~\ref{fig:ltfla_factorial}) and sum over an array
(Figure~\ref{fig:ltfla_sumarray}).

@@

The syntax is intended to resemble OCaml's, apart from the spurious
\ltfla{`!'}s found here and there (they are annotations to show that we can
use a value more than once). In Figure~\ref{fig:ltfla_sumarray}, we see
\ltfla{row} has type (\ltfla{'x arr}). More detailed explanations of
\emph{what and why} will be given in Chapter~\ref{chap:impl}, but for now, it
is enough to know it means we can only \emph{read} from \ltfla{row}, and not
write to it. If we did try to write to or free \ltfla{row}, we would get a
helpful error message, as shown in Figure~\ref{fig:ltfla_univ}.

\begin{figure}[tp]
    \begin{minted}[linenos, firstnumber=7, fontsize=\small]{ocaml}
let row = row[i] := x1 in (* or *) let () = free row in
(* Could not show equality:                                        *)
(*     z arr                                                       *)
(* with                                                            *)
(*     'x arr                                                      *)
(*                                                                 *)
(* Var 'x is universally quantified                                *)
(* Are you trying to write to/free/unshare an array you don't own? *)
(* In test/examples/sum_array.lt, at line: 7 and column: 19        *)
    \end{minted}
    \caption{Attempting to write to or free a read only array in
    LT4LA.}\label{fig:ltfla_univ}
\end{figure}

Now suppose we are trying to square a matrix, using a `dgemm' like BLAS
routine (called `simple\_dgemm'\footnote{For the purposes of this example, I
assume `simple\_dgemm' has type \ltfla{'x . 'x mat --o 'y .  'y mat --o z mat
--o ('x mat * 'y mat) * z mat}.} in Figures~\ref{fig:ltfla_square}
and~\ref{fig:ltfla_free}) which takes two read-only matrices and a third matrix
it can write to and performs $C := AB + C$. How would we use such a routine to
square a matrix, assuming $C=0$ initially?  Surely this would break linearity,
since $A$ would have to be passed in to the function twice, and we can only 
use non-!-annotated variables once?

To solve this, we use a special primitive called \ltfla{share} to produce
(more) read-only \emph{aliases} of any matrix. We then pass these into the
function, and it works as expected. Once the squared matrix has been obtained,
we may not want the original anymore, and thus decide to free it. Before we do
so, we must first \ltfla{unshare} any read-only aliases that exist, to get back
a single, read-write handle.

\begin{figure}[tp]
    \begin{minted}[linenos, fontsize=\small]{ocaml}
let (a1, a2) = share _ a in
let ((a1, a2), c) = simple_dgemm _ a1 _ a2 c in
let a = unshare _ a1 a2 in
let () = free a in c
    \end{minted}
    \caption{Squaring a matrix in LT4LA (assuming \ltfla{c} is initially
        0).}\label{fig:ltfla_square}
\end{figure}

If we tried to free one of the read-only aliases before or instead of
\ltfla{unshare}-ing them, then we would get the error shown in
Figure~\ref{fig:ltfla_free}. Briefly, a \ltfla{z arr} is a read-write array,
everything else (\ltfla{'x arr} or \ltfla{_ s arr}) is read-only. The types of
\ltfla{free}, \ltfla{share} and \ltfla{unshare} (see
Appendix~\ref{chap:primitives}) are set up so that you can only free something
when you have read-write access to it, guaranteed by linearity to be the only
name in scope with this capability (aliases can only be read-only). Conversely,
if we \emph{forgot} to \ltfla{free a}, we would also get \texttt{Variable a not
used} error.

@@

\begin{figure}[tp]
    \begin{minted}[linenos, fontsize=\small]{ocaml}
let (a1, a2) = share _ a in
let ((a1, a2), c) = simple_dgemm _ a1 _ a2 c in
let () = free a1 in c
(* Error:                                                        *)
(* Could not show equality:                                      *)
(*     z arr                                                     *)
(* with                                                          *)
(*     z s arr                                                   *)
(*                                                               *)
(* Could not show z and z s are equal.                           *)
(* Are you trying to write to/free an array before unsharing it? *)
(* In test.lt, at line: 3 and column: 17                         *)
    \end{minted}
    \caption{Attempting to free a read-only alias of matrix.}\label{fig:ltfla_free}
\end{figure}

@@

More generally, by annotating variables in the context with their usage (when
implementing a type checker for a linearly typed language), we can express the
rules of \emph{substructural} (including affine, relevant and ordered type
systems) under the more general framework of \emph{co-effects}~\cite{petricek}.

@@

I have given an \emph{intuitive} exposition of linear types with
\emph{fractional-capabilities}~\cite{boyland}, emphasising small, but
illustrative and practical code examples, leaving \emph{what} is going on and
\emph{why} it works as details for the next chapter. I have shown that it is
possible to solve the dichotomy of readability, ease of reasoning and safety
versus memory-efficiency by providing explicit and automatically checked
management of aliasing, read/write permissions, memory allocation, re-use and
deallocation with a \emph{type system}. This prevents a \textbf{whole class} of
errors that can occur with low-level languages \emph{at compile time}. For
example, \textbf{the types make it impossible to use `simple\_dgemm'
incorrectly}. I demonstrated these features using the context of linear algebra
libraries as as a specific example. In the next chapter, I will show that these
features can be further applied to the problem of \emph{matrix expression
compilation}. I will also explain how to express and implement them so that
they can be provided and used \emph{as a library}.

### 2018-05-28 write-up/conclusion.tex
Fix many small writing errors

In this thesis, I presented \textbf{linear types with fractional-capabilities}
as a type-based formalism for expressing \textbf{aliasing, read/write
permissions, memory allocation, re-use and deallocation}; I provided a detailed
description of each feature of the type-system, its intent and implementation
in Section~\ref{sec:core_lang}.

@@

I also used that formalism to implement a simple, yet effective matrix
expression `compiler' based on pattern-matching. I was able to use to that
compiler to write, a \textbf{non-trivial, yet readable} (equational and
algebraic) linear-algebra program (Kalman filter, as described in
Section~\ref{sec:matrix_exps}). I shared how, in doing so, I could use the
type-checking as a tool to \textbf{automatically highlight bugs} regarding
aliasing, read/write permissions and memory re-use of a reference
implementation (Section~\ref{sec:qual_benefits}) and the matrix expression
compiler that generated it (Section~\ref{sec:mat_exp_comp}).

@@

I showed that this implementation was \textbf{more memory-efficient} than one
written using a high-level linear-algebra libraries and performed just as
\textbf{predictably} as one written using a low-level linear-algebra library,
enabling a programmer to have the best of both worlds
(Chapter~\ref{chap:eval}). The type system ensured that the implementation was
\textbf{safe and explicit} with respect to aliasing, read/write permissions,
memory allocation, re-use and deallocation.

Lastly, I was able to provide all of this functionality as a \textbf{usable
OCaml library}, by generating \textbf{readable and not obviously safe} OCaml
code (1) which looks like it is written by a human/expert and (2) whose types
reflect its intensional behaviour and assumptions (interpreting types into
OCaml, Figure~\ref{fig:type_grammar}). This library, LT4LA, produces
\textbf{helpful type-errors} (Section~\ref{sec:prop_sol}), comes with an
interactive REPL and is documented and well-tested
(Section~\ref{sec:structure_lt4la}).

@@

LT4LA is a proof-of-concept design for linearly-typed DSL for writing
linear-algebra programs in a way that can work with existing languages and
libraries. It is made of up of the core language and its matrix expression
`compiler'; both could be extended in many ways.

@@

Although I use well-established type system features, it may be worthwhile to
formally state and verify my claimed safety properties.  The logical conclusion
of LT4LA's core language would be a language expressive enough to provide a
usable, linearly-typed interface to the unsafe parts of the Owl library for
OCaml, perhaps as a functor like its current, `Lazy' interface.  To do that,
the language implementation would need to support more than just 64-bit
floating-point numbers for elements (for example, 16-bit, 32-bit or complex),
polymorphism, a unified way of dealing with arrays and matrices (perhaps
through sub-typing). It may also be useful to explore localised linearity
\emph{inference}, to automate the reasoning I did manually when implementing a
memory-efficient Kalman filter.

Another, equally valid but opposing direction may be to decrease the amount of
expressivity, to bring it closer to Fortran and C so that it would be easier to
emit code in those languages. A full set of linearly-typed BLAS/LAPACK bindings
(to encode information in documentation at the type level) would be a good
first step, perhaps leading into more research on formal semantics for Fortran.

@@

    \item \textbf{As a typed IR} for matrix expression compilers. This in turn
        could enable
        \begin{itemize}
            \item existing matrix expression compilers to be less opaque about
                what resources they are consuming.
            \item open up opportunities for non-local sharing of temporary values
                with some intra-procedural analysis.
            \item allow the user to choose: use a matrix expression compiler when
                desired and drop down to a usable typed-IR for finer control,
                whilst still retaining safety guarantees.
        \end{itemize}

    \item \textbf{Formal verification of matrix expression compilers} by
        precisely specifying source and target languages.

    \item \textbf{Multi-stage programming} to use information only available
        at runtime in many situations (such as sizes, matrix properties,
        sparsity, control flow) can be effectively incorporated into code
        generated.

    \item \textbf{Dependent types} to have control over how resources can be
        used and split. In addition to formal verification, dependent types
        could be combined with linearity to express finer-grain conventions
        surrounding blocking, slicing and writing to \emph{parts} of the matrix
        instead of the whole. This is already prevalent with `dsymm' like BLAS
        routines which only read and write to the lower or upper triangle of a
        matrix.  This idea is inspired by Conor McBride's talk on writing to
        terminals with ``Space Monads''~\cite{space_monads}.

>>>>

In this thesis, I presented \textbf{linear types with fractional-capabilities}
as a type-based formalism for expressing \textbf{aliasing, read/write
permissions, memory allocation, re-use and deallocation}; I provided a detailed
description of each feature of the type system, its intent and implementation
in Section~\ref{sec:core_lang}.

@@

I also used that formalism to implement a simple, yet effective matrix
expression `compiler' based on pattern-matching. I was able to use to that
compiler to write, a \textbf{non-trivial, yet readable} (equational and
algebraic) linear algebra program (Kalman filter, as described in
Section~\ref{sec:matrix_exps}). I shared how, in doing so, I could use the
type-checking as a tool to \textbf{automatically highlight bugs} regarding
aliasing, read/write permissions and memory re-use of a reference
implementation (Section~\ref{sec:qual_benefits}) and the matrix expression
compiler that generated it (Section~\ref{sec:mat_exp_comp}).

@@

I showed that this implementation was \textbf{more memory-efficient} than one
written using a high-level library and performed just as
\textbf{predictably} as one written using a low-level library,
enabling a programmer to have the best of both worlds
(Chapter~\ref{chap:eval}). The type system ensured that the implementation was
\textbf{safe and explicit} with respect to aliasing, read/write permissions,
memory allocation, re-use and deallocation.

Lastly, I was able to provide all of this functionality as a \textbf{usable
OCaml library}, by generating \textbf{readable and not obviously safe} OCaml
code (1) which looks like it is written by an expert and (2) whose types
reflect its intensional behaviour and assumptions (interpreting types into
OCaml, Figure~\ref{fig:type_grammar}). This library, LT4LA, produces
\textbf{helpful type-errors} (Section~\ref{sec:prop_sol}), comes with an
interactive REPL and is documented and well-tested
(Section~\ref{sec:structure_lt4la}).

@@

LT4LA is a proof-of-concept design for linearly typed DSL for writing
linear-algebra programs in a way that can work with existing languages and
libraries. It is made of up of the core language and its matrix expression
`compiler'; both could be extended in many ways.

@@

Although I use well-established type system features, it may be worthwhile to
formally state and verify my claimed safety properties.  The logical conclusion
of LT4LA's core language would be a language expressive enough to provide a
usable, linearly typed interface to the unsafe parts of the Owl library for
OCaml, perhaps as a functor like its current, `Lazy' interface.  To do that,
the language implementation would need to support more than just 64-bit
floating-point numbers for elements (for example, 16-bit, 32-bit or complex),
polymorphism, a unified way of dealing with arrays and matrices (perhaps
through sub-typing). It may also be useful to explore localised linearity
\emph{inference}, to automate the reasoning I did manually when implementing a
memory-efficient Kalman filter (Section~\ref{sec:qual_benefits}).

Another, equally valid but opposing direction may be to decrease the amount of
expressivity, to bring it closer to Fortran and C so that it would be easier to
emit code in those languages. A full set of linearly typed BLAS/LAPACK bindings
(to encode information in documentation at the type level) would be a good
first step, perhaps leading into more research on formal semantics for Fortran.

@@

    \item \textbf{As a typed IR} for matrix expression compilers. This in turn
        could enable
        \begin{itemize}
            \item existing matrix expression compilers to be less opaque about
                what resources they are consuming.
            \item open up opportunities for non-local sharing of temporary values
                with some intra-procedural analysis.
            \item allow the user to choose: use a matrix expression compiler when
                desired and drop down to a usable typed IR for finer control,
                whilst still retaining safety guarantees.
        \end{itemize}

    \item \textbf{Formal verification of matrix expression compilers} by
        precisely specifying the intensional and extensional properties of the
        source and target languages.

    \item \textbf{Multi-stage programming} to use information only available at
        runtime (such as sizes, matrix properties, sparsity, control flow) for
        generating more specialised code.

    \item \textbf{Dependent types} to have control over how resources can be
        used and split. In addition to formal verification, dependent types
        could be combined with linearity to express finer-grain conventions
        surrounding blocking, slicing and writing to \emph{parts} of the matrix
        instead of the whole. This is already prevalent with `dsymm' like BLAS
        routines which only read the lower or upper triangle of a matrix.  This
        idea is inspired by Conor McBride's talk on writing to
        terminals with ``Space Monads''~\cite{space_monads}.
